The role of cognateness in non-native spoken word recognition

### 2022-04-27 manuscript/manuscript.Rmd
Restore workflow

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener's native language. For example, imagine a Spanish native, with no previous familiarity with any other language, listens for the first time to French may encounter the word /pɔʁt/ (*porte*), translation of *door* in French. The voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/ do not exist in Spanish. These discrepancies can nonetheless be perceived as allophonic variations of native phonemes [@best2001discrimination], and the non-native speech signal engages lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and therefore the recognition of its translation to Spanish /'pwer.ta/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /dɔː/ (*door*). This time, phonological similarity is of little help: /dɔː/ and /'pwer.ta/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations (which they lack, they are unfamiliar to English).

@@

Measuring the phonological similarity between two word forms is also substantially more challenging than measuring their orthographic similarity. Many studies use symbolic representations to compute the simiarity between word-forms of interest in either modality. Orthographic word-forms in alphabetical languages are represented by the strings of letters that make up the word (*radio*-*radio*); phonological word-forms are represented by their  phonological transcriptions (ɹeɪdiəʊ-raðjo, in IPA notation) [see @floccia2018introduction for example]. The similarity between each pair of word-forms is operationalised numerically by applying some algorithm that computes the edit distance between their symbolic representation. The Leveshtein distance is one of the most widely known algorithms: it computes the number of insertions, deletions or substitutions of the of the strings must go through in order to become identical to the other string [@levenshtein1966binary], and has been successfully used to estimate the similarity of cross-language word-pairs [@gooskens2004perceptive; @schepens2012distributions]. Using our previous examples, the Levenshtein distance between the orthographic word-forms *radio* (English) and *radio* (Spanish) is `r stringdist::stringdist("radio", "radio")`: both strings are identical. But the distance between the phonological transcriptions ɹeɪdiəʊ and raðjo is `r stringdist::stringdist("ɹeɪdiəʊ", "raðjo")`, which is the number of edits that separates both character strings. There is also an additional layer of complexity: the phonological transcriptions /ˈɹeɪdiˌəʊ/ and /ˈraðjo/ can also be transcribed in different ways depending on variation in pronunciation between speakers, and decisions on the transcriptions (e.g. whether to include elements of aspiration, nasalisation, etc.) can affect how accurate these symbolic representation capture all of the features that are present in the acoustic signal when either word is spoken. Therefore, any measure of similarity that uses phonological transcriptions as input may overestimate or underestimate their actual similarity. Taking into account that measuring phonological similarity is considerably more complex than measuring orthographic complexity it is no surprise that most literature on word recognition (or even on psycholinguistics) has been devoted to study orthography instead of phonology.

@@

Data collection took place from `r format(min(participants$date, na.rm = TRUE), "%B %dth, %Y")` to `r format(max(participants$date, na.rm = TRUE), "%B %dth, %Y")`. We collected data from `r length(unique(participants$participant))` participants (*Mean* = `r printnum(mean(participants$age, na.rm = TRUE))` years, *SD* = `r printnum(sd(participants$age, na.rm = TRUE))`, Range = `r printnum(min(participants$age, na.rm = TRUE))`-`r printnum(max(participants$age, na.rm = TRUE))`). `r length(unique(participants$participant[participants$group %in% c("ENG-CAT", "ENG-SPA")]))` participants were British English native speakers living in United Kingdom (`r length(unique(participants$participant[participants$group %in% c("ENG-CAT", "ENG-SPA") & participants$sex=="Female"]))` female), and `r length(unique(participants$participant[participants$group %in% c("SPA-CAT")]))` participants were Spanish native speakers living in Spain (`r length(unique(participants$participant[participants$group %in% c("SPA-CAT") & participants$sex=="Female"]))` female). Participants in UK were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits). Participants in Spain were contacted via announcements in Faculties, and were compensated 5€ or an Amazon voucher for the same value. Participants gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar (2020/9080/I). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection. We excluded data from participants that (a) self-rated their oral and/or written skills in a second or third language as higher than 4 in a 5-point scale (*n* = `r participants %>% filter(l2written > 4 | l2oral > 4) %>% nrow()`), (b) were diagnosed with a language (*n* = `r participants %>% filter(impairment) %>% nrow()`), or (c) did not contribute more than 80% of valid trials (*n* = `r participants %>% filter(invalid_participant_trials) %>% nrow()`).

@@

The experiment was implemented online using Psychopy/Pavlovia [@peirce2019psychopy2]. Participants accessed the study from a link provided by Prolific or SONA and completed the experiment from an internet browser (Chrome or Mozilla). After giving their consent for participating, participants answered a series of questions about their demographic status, their language background, and the set up they were using for completing the study. Then, participants completed the experimental task. Participants were informed that they would listen to a series of pre-recorded words in either Catalan or Spanish (English participants) or Catalan (Spanish participants). They were instructed to listen to each word, guess its meaning in English (English participants) or Spanish (Spanish participants), and type their answer as soon as possible. English participants were randomly assigned to the Catalan or Spanish lists. Participants in the Catalan list were presented with `r nrow(stimuli[stimuli$group=="ENG-CAT",])` trials, and participants in the Spanish list were presented with `r nrow(stimuli[stimuli$group=="ENG-SPA",])` trials. Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ">" symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN/ENTER key to confirm their answer and start and new trial. We excluded trials where the participant's answer was not a valid word in the target language or no answer was given. Trials in which the response was mistyped by only one character were counted as correct (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct word. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(responses)` valid trials (`r nrow(responses[responses$group %in% c("ENG-CAT", "SPA-CAT"),])` in Catalan, `r nrow(responses[responses$group %in% c("ENG-SPA"),])` in Spanish). The task took approximately 15 minutes to be completed.

@@

We arranged two lists of input words to be presented to participants in the auditory modality: one in Catalan (to be presented to English and Spanish natives) and one in Spanish (to be presented to English natives only). Words in the Catalan list were `r printnum(lengths$eng_cat["ipa1_mean"])` phonemes long on average (*SD* = `r printnum(lengths$eng_cat["ipa1_sd"])`, Range = `r printnum(lengths$eng_cat["ipa1_min"], digits = 0)`-`r printnum(lengths$eng_cat["ipa1_max"], digits = 0)`). Words in the Spanish list were `r printnum(lengths$eng_spa["ipa1_mean"])` phonemes long on average (*SD* = `r printnum(lengths$eng_spa["ipa1_sd"])`, Range = `r printnum(lengths$eng_spa["ipa1_min"], digits = 0)`-`r printnum(lengths$eng_spa["ipa1_max"], digits = 0)`). Participants listened to one audio file in each trial, each containing a single word presented in isolation. The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the audio files was `r printnum(mean(stimuli$duration, na.rm = TRUE))` (*SD* = `r printnum(sd(stimuli$duration, na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration, na.rm = TRUE))`-`r printnum(max(stimuli$duration, na.rm = TRUE))`). The average duration of the Catalan audio files was `r printnum(mean(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`), and the average duration of the Spanish audio files was `r printnum(mean(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`).

@@

Responses given by English participants for Catalan presented words were `r printnum(lengths$eng_cat["word2_mean"])` characters long on average (*SD* = `r printnum(lengths$eng_cat["word2_sd"])`, Range = `r printnum(lengths$eng_cat["word2_min"], digits = 0)`-`r printnum(lengths$eng_cat["word2_max"], digits = 0)`), while their translations to Spani responses given by Spanish participants for Catalan words were `r printnum(lengths$spa_cat["word2_mean"])` characters long on average (*SD* = `r printnum(lengths$spa_cat["word2_sd"])`, Range = `r printnum(lengths$eng_cat["word2_min"], digits = 0)`-`r printnum(lengths$eng_cat["word2_max"], digits = 0)`). Responses tgiven by English participants for Spanish words were `r printnum(lengths$eng_spa["word2_mean"])` characters long on average (*SD* = `r printnum(lengths$eng_spa["word2_sd"])`, Range = `r printnum(lengths$eng_spa["word2_min"], digits = 0)`-`r printnum(lengths$eng_spa["word2_max"], digits = 0)`).

@@

pthn_example <- clearpond_path %>% 
  map(read.csv) %>% 
  bind_rows(.id = "group") %>% 
  clean_names() %>% 
  filter(
    word %in% c("book", "sock"),
    group=="ENG-CAT"
  ) %>% 
  select(word, pthn)

@@

d <- stimuli %>% 
  mutate(
    group = case_when(
      group=="ENG-CAT" ~ "cat-ENG",
      group=="ENG-SPA" ~ "spa-ENG",
      group=="SPA-CAT" ~ "cat-SPA"
    ),
    group = factor(
      group,
      levels = c(
        "cat-ENG",
        "spa-ENG", 
        "cat-SPA"
      )
    )
  )

@@

We measured the phonological similarity between presented words and their target translation by computing the Levenshtein similarity between their IPA transcriptions using the `stringsim` function of the stringdist R package [@van2014stringdist]. This function computes the inverse of the Levenshtein distance between two character strings as a proportion. First, it computes the edit distance between two character strings (in this case, phoneme transcriptions) by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. This measure is then divided by the maximum distance (according to the length of the longest string) and then subtracted from 1. The result is a percentage score, with 0% indicating no similarity between the two strings, and 100% indicating a perfect match between the two strings. We computed this similarity measure (refred simply as Levenshtein, from now on) for every translation pair in our stimuli lists. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r percent(stringsim("teɪbəl", "mesa"))` similarity, while the *train* (trEIn)-*tren* (tɾEn) translation pair had a `r percent(stringsim("trEIn", "tɾEn"))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

str_repl <- c(
  "Intercept" = "Intercept",
  "frequency_zipf_std" = "Frequency\n[+1 SD]",
  "pthn_std:lv_std" = "PTHN \u00d7 Levenshtein", 
  "lv_std:groupENGvs.SPA" = "Levenshtein \u00d7 Group 1",
  "lv_std:groupENGMCATvs.ENGMSPA" = "Levenshtein \u00d7 Group 2",
  "lv_std" = "Levenshtein\n[+1 SD]",
  "pthn_std" = "PTHN\n[+1 SD]",
  "groupENGvs.SPA" =  "Group 1\n[cat/spa-ENG vs. cat-SPA]",
  "groupENGMCATvs.ENGMSPA" = "Group 2\n[cat-ENG vs. spa-ENG]"
)

@@

post_sd <- post_draws %>% 
  filter(group=="Participants") %>% 
  ggplot() +
  aes(.value, fct_rev(.variable_name)) +
  stat_slab(
    aes(
      fill = stat(
        cut_cdf_qi(
          cdf, 
          .width = c(.5, .8, .95),
          labels = percent_format()
        )
      )
    )
  ) +
  stat_pointinterval(
    .width = 0, 
    point_size = 1
  ) +
  scale_fill_manual(
    values = c("#ff2976", "#f2a7c2", "#fcd9e6"), 
    na.translate = FALSE
  ) +
  scale_x_continuous(
    labels = percent, 
    limits = c(0, 1)
  ) +
  labs(
    x = "SDs in P(Correct)",
    title = "By participant", 
    fill = "CrI"
  ) +
  theme(
    axis.text.x = element_blank()
  ) +

@@

# Pearson correlations between random effects
corr_mat <- VarCorr(fit_6, summary = TRUE)$participant$cor 
corr_names <- rownames(VarCorr(fit_6, summary = TRUE)$participant$cor)

@@

h_frequency <- get_bf("frequency_zipf_std = 0")
h_pthn <- get_bf("pthn_std = 0")
h_lv <- get_bf("pthn_std = 0")
h_pthn_lv <- get_bf("pthn_std:lv_std = 0")
h_group1 <- get_bf("groupENGvs.SPA = 0")
h_group2 <- get_bf("groupENGMCATvs.ENGMSPA = 0")
h_lv_group1 <- get_bf("lv_std:groupENGvs.SPA = 0")
h_lv_group2 <- get_bf("lv_std:groupENGMCATvs.ENGMSPA = 0")

@@

word_example <- ranef(fit_6)$word[,,"Intercept"] %>% 
  as.data.frame() %>% 
  rownames_to_column("word") %>% 
  arrange(-Estimate) %>% 
  clean_names() %>% 
  mutate_if(is.numeric, inv_logit_scaled) %>% 
  slice(1, n()) %>% 
  left_join(
    distinct(
      stimuli,
      word1, 
      ipa1), 
    by = c("word" = "word1")) %>% 
  split(.$word)

@@

corr_mat <- VarCorr(fit_6, summary = TRUE)$participant$cor 
corr_names <- rownames(VarCorr(fit_6, summary = TRUE)$participant$cor)

@@

h_cor_intercept_frequency <- get_bf(
  "participant__Intercept__frequency_zipf_std = 0", 
  class = "cor"
)
h_cor_frequency_zipf_lv_group1 <- get_bf(
  "participant__frequency_zipf_std__lv_std:groupENGvs.SPA = 0", 
  class = "cor"
)
h_cor_intercept_pthn <- get_bf(
  "participant__Intercept__pthn_std = 0",
  class = "cor"
)
```

@@

Spanish natives translating Catalan words were `r percent(coefs$group_en_gvs_spa["estimate"], accuracy = 0.01)` more accurate than English natives translating Spanish and Catalan words (95% *CrI* = [`r percent(coefs$group_en_gvs_spa["q2_5"], accuracy = 0.01)`, `r percent(coefs$group_en_gvs_spa["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_group1)`). The interaction between this contrast and Levenshtein suggests a moderate impact of phonological similarity on the difference between Spanish and English participants: this difference increased in `r percent(coefs$lv_std_group_en_gvs_spa["estimate"], accuracy = 0.01)` for every one *SD* increment in Levenshtein similarity (95% *CrI* = [`r percent(coefs$lv_std_group_en_gvs_spa["q2_5"], accuracy = 0.01)`, `r percent(coefs$lv_std_group_en_gvs_spa["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_lv_group1)`). English natives translating Catalan words were `r percent(abs(coefs$group_engmca_tvs_engmspa["estimate"]), accuracy = 0.01)` more accurate than those translating Spanish words (95% *CrI* = [`r percent(coefs$group_engmca_tvs_engmspa["q2_5"], accuracy = 0.01)`, `r percent(coefs$group_engmca_tvs_engmspa["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_group2)`). For every one *SD* increment in Levenshtein similarity, the effect of presented language increased in `r percent(abs(coefs$lv_std_group_engmca_tvs_engmspa["estimate"]), accuracy = 0.01)` (95% *CrI* = [`r percent(coefs$lv_std_group_engmca_tvs_engmspa["q2_5"], accuracy = 0.01)`, `r percent(coefs$lv_std_group_engmca_tvs_engmspa["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_lv_group1)`).

>>>>

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener's native language. For example, imagine a Spanish native, with no previous familiarity with any other language, listens for the first time to French may encounter the word p`r knitr::asis_output("\U0254")``r knitr::asis_output("\U0281")`t/ (*porte*), translation of *door* in French. The voiced uvular fricative consonant /`r knitr::asis_output("\U0281")`/ and the open-mid back rounded vowel /`r knitr::asis_output("\U0254")`/ do not exist in Spanish. These discrepancies can nonetheless be perceived as allophonic variations of native phonemes [@best2001discrimination], and the non-native speech signal engages lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and therefore the recognition of its translation to Spanish /'pwer.ta/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /d`r knitr::asis_output("\U0254")`ː/ (*door*). This time, phonological similarity is of little help: /d`r knitr::asis_output("\U0254")`ː/ and /'pwer.ta/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations (which they lack, they are unfamiliar to English).

@@

Measuring the phonological similarity between two word forms is also substantially more challenging than measuring their orthographic similarity. Many studies use symbolic representations to compute the simiarity between word-forms of interest in either modality. Orthographic word-forms in alphabetical languages are represented by the strings of letters that make up the word (*radio*-*radio*); phonological word-forms are represented by their  phonological transcriptions (`r knitr::asis_output("\U0279")`eɪdiəʊ-raðjo, in IPA notation) [see @floccia2018introduction for example]. The similarity between each pair of word-forms is operationalised numerically by applying some algorithm that computes the edit distance between their symbolic representation. The Leveshtein distance is one of the most widely known algorithms: it computes the number of insertions, deletions or substitutions of the of the strings must go through in order to become identical to the other string [@levenshtein1966binary], and has been successfully used to estimate the similarity of cross-language word-pairs [@gooskens2004perceptive; @schepens2012distributions]. Using our previous examples, the Levenshtein distance between the orthographic word-forms *radio* (English) and *radio* (Spanish) is `r stringdist::stringdist("radio", "radio", useBytes = TRUE)`: both strings are identical. But the distance between the phonological transcriptions `r knitr::asis_output("\U0279")`eɪdiə`r knitr::asis_output("\U028A")` and ra`r knitr::asis_output("\U00F0")`jo is 7, which is the number of edits that separates both character strings. There is also an additional layer of complexity: the phonological transcriptions /ˈ`r knitr::asis_output("\U0279")`eɪdiˌə`r knitr::asis_output("\U028A")`/ and /ˈra`r knitr::asis_output("\U00F0")`jo/ can also be transcribed in different ways depending on variation in pronunciation between speakers, and decisions on the transcriptions (e.g. whether to include elements of aspiration, nasalisation, etc.) can affect how accurate these symbolic representation capture all of the features that are present in the acoustic signal when either word is spoken. Therefore, any measure of similarity that uses phonological transcriptions as input may overestimate or underestimate their actual similarity. Taking into account that measuring phonological similarity is considerably more complex than measuring orthographic complexity it is no surprise that most literature on word recognition (or even on psycholinguistics) has been devoted to study orthography instead of phonology.

@@

Data collection took place from `r format(min(participants$date, na.rm = TRUE), "%B %dth, %Y")` to `r format(max(participants$date, na.rm = TRUE), "%B %dth, %Y")`. We collected data from `r length(unique(participants$participant_id))` participants (*Mean* = `r printnum(mean(participants$age, na.rm = TRUE))` years, *SD* = `r printnum(sd(participants$age, na.rm = TRUE))`, Range = `r printnum(min(participants$age, na.rm = TRUE))`-`r printnum(max(participants$age, na.rm = TRUE))`). `r length(unique(participants$participant_id[participants$group %in% c("cat-ENG", "spa-ENG")]))` participants were British English native speakers living in United Kingdom (`r length(unique(participants$participant_id[participants$group %in% c("cat-ENG", "spa-ENG") & participants$sex=="Female"]))` female), and `r length(unique(participants$participant_id[participants$group %in% c("cat-SPA")]))` participants were Spanish native speakers living in Spain (`r length(unique(participants$participant_id[participants$group %in% c("cat-SPA") & participants$sex=="Female"]))` female). Participants in UK were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits). Participants in Spain were contacted via announcements in Faculties, and were compensated 5€ or an Amazon voucher for the same value. Participants gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar (2020/9080/I). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection. We excluded data from participants that (a) self-rated their oral and/or written skills in a second or third language as higher than 4 in a 5-point scale (*n* = `r participants %>% filter(l2written > 4 | l2oral > 4) %>% nrow()`), (b) were diagnosed with a language (*n* = `r participants %>% filter(impairment) %>% nrow()`), or (c) did not contribute more than 80% of valid trials (*n* = `r participants %>% filter(invalid_participant_trials) %>% nrow()`).

@@

The experiment was implemented online using Psychopy/Pavlovia [@peirce2019psychopy2]. Participants accessed the study from a link provided by Prolific or SONA and completed the experiment from an internet browser (Chrome or Mozilla). After giving their consent for participating, participants answered a series of questions about their demographic status, their language background, and the set up they were using for completing the study. Then, participants completed the experimental task. Participants were informed that they would listen to a series of pre-recorded words in either Catalan or Spanish (English participants) or Catalan (Spanish participants). They were instructed to listen to each word, guess its meaning in English (English participants) or Spanish (Spanish participants), and type their answer as soon as possible. English participants were randomly assigned to the Catalan or Spanish lists. Participants in the Catalan list were presented with `r nrow(stimuli[stimuli$group=="cat-ENG",])` trials, and participants in the Spanish list were presented with `r nrow(stimuli[stimuli$group=="spa-ENG",])` trials. Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ">" symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN/ENTER key to confirm their answer and start and new trial. We excluded trials where the participant's answer was not a valid word in the target language or no answer was given. Trials in which the response was mistyped by only one character were counted as correct (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct word. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(responses)` valid trials (`r nrow(responses[responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(responses[responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

@@

We arranged two lists of input words to be presented to participants in the auditory modality: one in Catalan (to be presented to English and Spanish natives) and one in Spanish (to be presented to English natives only). Words in the Catalan list were `r printnum(lengths$cat_eng["ipa1_mean"])` phonemes long on average (*SD* = `r printnum(lengths$cat_eng["ipa1_sd"])`, Range = `r printnum(lengths$cat_eng["ipa1_min"], digits = 0)`-`r printnum(lengths$cat_eng["ipa1_max"], digits = 0)`). Words in the Spanish list were `r printnum(lengths$spa_eng["ipa1_mean"])` phonemes long on average (*SD* = `r printnum(lengths$spa_eng["ipa1_sd"])`, Range = `r printnum(lengths$spa_eng["ipa1_min"], digits = 0)`-`r printnum(lengths$spa_eng["ipa1_max"], digits = 0)`). Participants listened to one audio file in each trial, each containing a single word presented in isolation. The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the audio files was `r printnum(mean(stimuli$duration, na.rm = TRUE))` (*SD* = `r printnum(sd(stimuli$duration, na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration, na.rm = TRUE))`-`r printnum(max(stimuli$duration, na.rm = TRUE))`). The average duration of the Catalan audio files was `r printnum(mean(stimuli$duration[stimuli$group %in% c("cat-ENG")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("cat-ENG")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("cat-ENG")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("cat-ENG")], na.rm = TRUE))`), and the average duration of the Spanish audio files was `r printnum(mean(stimuli$duration[stimuli$group %in% c("spa-ENG")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("spa-ENG")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("spa-ENG")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("spa-ENG")], na.rm = TRUE))`).

@@

Responses given by English participants for Catalan presented words were `r printnum(lengths$cat_eng["word2_mean"])` characters long on average (*SD* = `r printnum(lengths$cat_eng["word2_sd"])`, Range = `r printnum(lengths$cat_eng["word2_min"], digits = 0)`-`r printnum(lengths$cat_eng["word2_max"], digits = 0)`), while their translations to Spanish responses given by Spanish participants for Catalan words were `r printnum(lengths$cat_spa["word2_mean"])` characters long on average (*SD* = `r printnum(lengths$cat_spa["word2_sd"])`, Range = `r printnum(lengths$cat_spa["word2_min"], digits = 0)`-`r printnum(lengths$cat_spa["word2_max"], digits = 0)`). Responses given by English participants for Spanish words were `r printnum(lengths$spa_eng["word2_mean"])` characters long on average (*SD* = `r printnum(lengths$spa_eng["word2_sd"])`, Range = `r printnum(lengths$spa_eng["word2_min"], digits = 0)`-`r printnum(lengths$spa_eng["word2_max"], digits = 0)`).

@@

pthn_example <- clearpond_path %>% 
  map(read.csv) %>% 
  bind_rows(.id = "group") %>% 
  clean_names() %>% 
  filter(
    word %in% c("book", "sock"),
    group=="cat-ENG"
  ) %>% 
  select(word, pthn)

@@

d <- stimuli %>% 
  mutate(
    group = factor(
      group,
      levels = c(
        "cat-ENG",
        "spa-ENG", 
        "cat-SPA"
      )
    )
  )

@@

We measured the phonological similarity between presented words and their target translation by computing the Levenshtein similarity between their IPA transcriptions using the `stringsim` function of the stringdist R package [@van2014stringdist]. This function computes the inverse of the Levenshtein distance between two character strings as a proportion. First, it computes the edit distance between two character strings (in this case, phoneme transcriptions) by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. This measure is then divided by the maximum distance (according to the length of the longest string) and then subtracted from 1. The result is a percentage score, with 0% indicating no similarity between the two strings, and 100% indicating a perfect match between the two strings. We computed this similarity measure (refred simply as Levenshtein, from now on) for every translation pair in our stimuli lists. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (trEIn)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

str_repl <- c(
  "Intercept" = "Intercept",
  "frequency_zipf_std" = "Frequency\n[+1 SD]",
  "pthn_std:lv_std" = "PTHN \u00d7 Levenshtein", 
  "lv_std:groupeng_spa" = "Levenshtein \u00d7 Group 1",
  "lv_std:groupcateng_spaeng" = "Levenshtein \u00d7 Group 2",
  "lv_std" = "Levenshtein\n[+1 SD]",
  "pthn_std" = "PTHN\n[+1 SD]",
  "groupeng_spa" =  "Group 1\n[cat/spa-ENG vs. cat-SPA]",
  "groupcatengspaeng" = "Group 2\n[cat-ENG vs. spa-ENG]"
)

@@

post_sd <- post_sd_data %>% 
  filter(group=="Participants") %>% 
  ggplot() +
  aes(
    x = .value, 
    y = fct_rev(.variable_name)
  ) +
  stat_slab(
    aes(
      fill = stat(
        cut_cdf_qi(
          cdf, 
          .width = c(.5, .8, .95),
          labels = percent_format()
        )
      )
    )
  ) +
  stat_pointinterval(
    .width = 0, 
    point_size = 1
  ) +
  scale_fill_manual(
    values = c("#ff2976", "#f2a7c2", "#fcd9e6"), 
    na.translate = FALSE
  ) +
  scale_x_continuous(
    labels = percent, 
    limits = c(0, 1)
  ) +
  labs(
    x = "SDs in P(Correct)",
    title = "By participant", 
    fill = "CrI"
  ) +
  theme(
    axis.text.x = element_blank()
  ) +

@@

# pearson correlations between random effects
corr_mat <- VarCorr(fit_6, summary = TRUE)$participant$cor 
corr_names <- rownames(VarCorr(fit_6, summary = TRUE)$participant$cor)

@@

h_frequency <- get_bf("frequency_zipf_std = 0")
h_pthn <- get_bf("pthn_std = 0")
h_lv <- get_bf("pthn_std = 0")
h_pthn_lv <- get_bf("pthn_std:lv_std = 0")
h_group1 <- get_bf("groupeng_spa = 0")
h_group2 <- get_bf("groupcateng_spaeng = 0")
h_lv_group1 <- get_bf("lv_std:groupeng_spa = 0")
h_lv_group2 <- get_bf("lv_std:groupcateng_spaeng = 0")

@@

word_example <- ranef(fit_6)$translation_id[,,"Intercept"] %>% 
  as.data.frame() %>% 
  rownames_to_column("word") %>% 
  arrange(-Estimate) %>% 
  clean_names() %>% 
  mutate_if(is.numeric, inv_logit_scaled) %>% 
  slice(1, n()) %>% 
  left_join(
    distinct(
      stimuli,
      word1, 
      ipa1), 
    by = c("word" = "word1")) %>% 
  split(.$word)

@@

corr_mat <- VarCorr(fit_6, summary = TRUE)$participant_id$cor 
corr_names <- rownames(VarCorr(fit_6, summary = TRUE)$participant_id$cor)

@@

h_cor_intercept_frequency <- get_bf(
  "participant_id__Intercept__frequency_zipf_std = 0", 
  class = "cor"
)
h_cor_frequency_zipf_lv_group1 <- get_bf(
  "participant_id__frequency_zipf_std__lv_std:groupeng_spa = 0", 
  class = "cor"
)
h_cor_intercept_pthn <- get_bf(
  "participant_id__Intercept__pthn_std = 0",
  class = "cor"
)
```

@@

Spanish natives translating Catalan words were `r percent(coefs$groupeng_spa["estimate"], accuracy = 0.01)` more accurate than English natives translating Spanish and Catalan words (95% *CrI* = [`r percent(coefs$groupeng_spa["q2_5"], accuracy = 0.01)`, `r percent(coefs$groupeng_spa["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_group1)`). The interaction between this contrast and Levenshtein suggests a moderate impact of phonological similarity on the difference between Spanish and English participants: this difference increased in `r percent(coefs$lv_std_groupeng_spa["estimate"], accuracy = 0.01)` for every one *SD* increment in Levenshtein similarity (95% *CrI* = [`r percent(coefs$lv_std_groupeng_spa["q2_5"], accuracy = 0.01)`, `r percent(coefs$lv_std_groupeng_spa["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_lv_group1)`). English natives translating Catalan words were `r percent(abs(coefs$groupcateng_spaeng["estimate"]), accuracy = 0.01)` more accurate than those translating Spanish words (95% *CrI* = [`r percent(coefs$groupcateng_spaeng["q2_5"], accuracy = 0.01)`, `r percent(coefs$groupcateng_spaeng["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_group2)`). For every one *SD* increment in Levenshtein similarity, the effect of presented language increased in `r percent(abs(coefs$lv_std_groupcateng_spaeng["estimate"]), accuracy = 0.01)` (95% *CrI* = [`r percent(coefs$lv_std_groupcateng_spaeng["q2_5"], accuracy = 0.01)`, `r percent(coefs$lv_std_groupcateng_spaeng["q97_5"], accuracy = 0.01)`], $BF_{\beta \neq 0}$ = `r printnum(h_lv_group1)`).
